Two Decades of Smalltalk VM Development

### 2018-08-18 paper.tex
Fixed a typo in Acknowledgements
typos

One of the key design decisions in \openSmalltalkVM was to keep the core components (i.e. the interpreter, the JIT and the memory manager) written in Slang and not in C/C++.  By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution. This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.\footnote{Arriving at the startup that first funded the OpenSmalltalk-VM, the first author had considerable experience maintaining the 2nd generation Deutsch Schiffmann VM written in C, and was sceptical that the Squeak Simulator would be viable for implementing a JIT.  The startup wanted incremental development so as to reap value early and reduce risk, so it asked first for an interpreter using context-to-stack mapping. Writing this initial deliverable within the simulator was so much more productive and pleasurable than his previous work in C that he happily decided to stay with the simulator and invest in the necessary machine code simulation infrastructure.}

@@

Debugging graph rewriting garbage collectors can be hard; bugs leave the heap in a scrambled state, and figuring out what happened when one is left with a corrupted graph, often too broken to traverse and make sense of, or apply tools to, is often effectively impossible.  But because the simulation is simply an object graph it is easy to copy the entire simulation. Therefore the simulator has been extended to by default take a copy of itself on every scavenge or old space reclamation, and to run the collection in the copy, performing a leak check afterwards (if enabled).  Hence if a bug occurs or validation fails it will fail in the copy while the original remains as is, yet to be corrupted.  We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging.  One throws the copy off the cliff and if it gets to the bottom without incident it is descarded and the simulation continues in the original.  This approach was key in rapid development of the new memory manager that supports both 32 and 64 bits \cite{Forwarders}.

@@

%% Acknowledgments
%%\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
\section*{Acknowledgements}
The authors would like to thank Ryan Macnak who write the MIPSLE sprocessor simulator and MIPSLE back end.
%  This material is based upon work supported by the
%  \grantsponsor{GS100000001}{National Science
%    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%  conclusions or recommendations expressed in this material are those
%  of the author and do not necessarily reflect the views of the
%  National Science Foundation.
%\end{acks}

>>>>

One of the key design decisions in \openSmalltalkVM was to keep the core components (i.e. the interpreter, the JIT and the memory manager) written in Slang and not in C/C++.  By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution. This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development\footnote{Arriving at the startup that first funded the OpenSmalltalk-VM, the first author had considerable experience maintaining the 2nd generation Deutsch Schiffmann VM written in C, and was sceptical that the Squeak Simulator would be viable for implementing a JIT.  The startup wanted incremental development so as to reap value early and reduce risk, so it asked first for an interpreter using context-to-stack mapping. Writing this initial deliverable within the simulator was so much more productive and pleasurable than his previous work in C that he happily decided to stay with the simulator and invest in the necessary machine code simulation infrastructure.}.

@@

Debugging graph rewriting garbage collectors can be hard; bugs leave the heap in a scrambled state, and figuring out what happened when one is left with a corrupted graph, often too broken to traverse and make sense of, or apply tools to, is often effectively impossible.  But because the simulation is simply an object graph it is easy to copy the entire simulation. Therefore the simulator has been extended to by default take a copy of itself on every scavenge or old space reclamation, and to run the collection in the copy, performing a leak check afterwards (if enabled).  Hence if a bug occurs or validation fails it will fail in the copy while the original remains as is, yet to be corrupted.  We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging%\footnote{http://www.romainbrasier.fr/404/}
.  One throws the copy off the cliff and if it gets to the bottom without incident it is descarded and the simulation continues in the original.  This approach was key in rapid development of the new memory manager that supports both 32 and 64 bits \cite{Forwarders}.

@@

%% Acknowledgments
%%\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
\section*{Acknowledgements}
The authors would like to thank Ryan Macnak who write the MIPSLE processor simulator and MIPSLE back end.
%  This material is based upon work supported by the
%  \grantsponsor{GS100000001}{National Science
%    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%  conclusions or recommendations expressed in this material are those
%  of the author and do not necessarily reflect the views of the
%  National Science Foundation.
%\end{acks}
